---
title: "wateRuse"
author: "Molly, Cheryl, etc..."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    fig_height: 5
    fig_width: 5
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{wateRuse Introduction}
  \usepackage[utf8]{inputenc}
---

# Introduction

The `wateRuse` package includes basic data loading, quality assurance / quality control (QA/QC), plotting, and mapping functions for USGS water use data. The package is designed to standarize and streamline USGS state and regional water use (WU) specialists' workflows. 

Water Science Center WU specialists can import current working AWUDS area files for their state as an export file (.xlsx format) and national data files (.csv format) which includes historical approved datasets. `wateRuse` can be used to merge the working export and national data files. The merged file is the basis for subsequent QA/QC. 

Regional water use specialists can import a national dump file that will include working and approved data. 

Users can display data based on time series or areas. Time series plots are for annual water use values for multiple years. Areas may include counties, HUC-08, aquifer, and state values. Users can compare different data elements by area and year and produce scatter plots, bar charts, line plots, and maps. 



## Importing data


## Data analysis tools
This package provides basic data visualization tools enable a user to graphically display water use data for different types of water uses by different areas. 

### Compare two years
The `compare_two_years` function enables the user to examine data for different data elements or areas between two years. This plot can be produced as a scatterplot. A comparison between two years for a single data element can be graphed for an entire collection of areas (e.g., all counties in a state or all states in the US) or for a subset of areas (e.g., top twenty counties in a state or selected states in a region). `areas` can be counties, HUC8s or aquifers. 

Users can also request multiple data elements, which results in one plot per data element. When areas are subsetted for a data element, a single graph is produced. 


FAQ: My compare two years plot doesn't have any data plotted, what's wrong? Answer: There may be data missing for the years selected. 


```{r message=FALSE}
library(wateRuse)
w.use <- wUseSample
areas <- "10" # NA uses all areas
area.column <- "STATECODE"
year.x.y <- c(2005,2010)
compare_two_years(w.use, "PS.TOPop", year.x.y,area.column)


```


### Ranked data element tables
In RStudio, users can sort data elements by loading data and using the RStudio viewer to sort by data element values.

### Time series plots


  
```{r message=FALSE}
areas <- c("Kent County","Sussex County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","PS.SWPop")
time_series_data(w.use, data.elements, area.column = area.column,areas = areas)

```
  
 

# Disclaimer
Software created by USGS employees along with contractors and grantees (unless specific stipulations are made in a contract or grant award) are to be released as Public Domain and free of copyright or license. Contributions of software components such as specific algorithms to existing software licensed through a third party are encouraged, but those contributions should be annotated as freely available in the Public Domain wherever possible. If USGS software uses existing licensed components, those licenses must be adhered to and redistributed.

Although this software has been used by the U.S. Geological Survey (USGS), no warranty, expressed or implied, is made by the USGS or the U.S. Government as to accuracy and functionality, nor shall the fact of distribution constitute any such warranty, and no responsibility is assumed by the USGS in connection therewith.