---
title: "wateRuse"
author: "Molly, Cheryl, etc..."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    fig_height: 5
    fig_width: 5
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{wateRuse Introduction}
  \usepackage[utf8]{inputenc}
---

# Introduction

The `wateRuse` package includes basic water-use data loading, plotting, and mapping functions that can be used to illustrate data, or perform QA/QC analysis on water-use data during compilation efforts.  The package is designed to standarize and streamline USGS Water Science Center (WSC) and regional water use (WU) specialists' workflows. 

WSC WU specialists can use current working AWUDS area files for their state (export file in .xlsx format) and national data files (.csv format), which includes historical approved datasets. `wateRuse` can be used to merge the working export and national approved data files. The merged file is the basis for subsequent plotting and analysis. 

Regional water use specialists can import a national dump file that will include working and approved data. 

Users can display water-use data for various data elements based on time (year) or areas (county, HUC8, aquifer, state). Time series plots us water-use values for a data element and  multiple years. A data element would represent the amount of water that is used from a specific source (groundwater, surface water) for a given area (ie county) and purpose (public supply).  Users can compare different data elements by area and year and produce scatter plots, bar charts, line graphs, and choropleth maps. 



## Importing data
A WSC user will navigate to a directory with a national best available dataset that contains all approved data that is viewable by all water-use users.  In addition to this, a WSC user will execute an AWUDS export function to produce an excel file (using default AWUDS naming convention).   These data files will be merged into one.  

A Regional user will navigate to a directory with all national data, including best available and current working data for all States.  

The ability to see data, whether national approved or state working data, will depend on the user type.  WSC users will always see their own working unapproved data.  Regional users will see a national dataset with approved and working data for review purposes. 

## Data analysis tools
This package provides basic data visualization tools that provide graphical displays of water-use data for different types of water uses by area and years. 

### Compare two years
The `compare_two_years` function enables the user to examine data by  area (county, HUC8, aquifer) for two years. This plot is produced as a scatterplot with a 1:1 line of equivalence to aid comparison. A comparison between two years for a single data element is graphed for an entire user-selected collection of areas (e.g., all counties in a state or all states in the US) or for a subset of areas (e.g., top twenty counties in a state or selected states in a region).

Users should be aware of what data elements exist for different years. For example, if a data element exists in one year but not both of the selected years, the plot will not be produced.
FAQ: My compare two years plot doesn't have any data plotted, what's wrong? Answer: There may be data missing for the years selected. 


```{r message=FALSE}
w.use <- wUseSample
data.elements <- c("PS.TOPop", "PS.SWPop")
areas <- "15" # NA uses all areas
area.column <- "STATECODE"
year.x.y <- c(2005,2010)
compare_two_years(w.use, "PS.TOPop", year.x.y, area.column, areas)

```

Users can specify multiple data elements which results in a stacked group of plots, with one plot per data element. 

```{r message=FALSE}
w.use <- wUseSample
data.elements <- c("PS.TOPop", "PS.SWPop")
areas <- "10" # NA uses all areas
area.column <- "STATECODE"
year.x.y <- c(2005,2010)
compare_two_years(w.use, data.elements, year.x.y, area.column, areas)

```

Multiple areas plot on the same plot.

```{r message=FALSE}
w.use <- wUseSample
data.elements <- "PS.TOPop"
areas <- c("10","15") # NA uses all areas
area.column <- "STATECODE"
year.x.y <- c(2005,2010)
compare_two_years(w.use, data.elements, year.x.y, area.column, areas)

```


### Compare two elements 

This function produces a scatterplot for comparison of two data elements for the same year or multiple years, for example, the number of irrigated acres and total irrigated withdrawals for a state for a given year.  Users can specify multiple years which results in a stacked group of plots, with one plot per year.  

```{r message=FALSE}
w.use <- wUseSample
data.elements.x.y <- c("TP.TotPop", "PS.WSWFr")
areas <- "15" # NA uses all areas
area.column <- "STATECODE"
years <- c(2000, 2005, 2010)
compare_two_elements(w.use, data.elements.x.y, years, area.column, areas)

```

Multiple areas are plotted together on the same plot.

```{r message=FALSE}
w.use <- wUseSample
compare_two_elements(w.use, c("TP.TotPop","PS.WSWFr"), "2010", "STATECODE", c("10","15"))
```

### Multi-Elements
Creates time series plots for one or more elements. Multiple data elements for a single area of interest (e.g. county) appear on the same plot. If more than one area is specified, multiple stacked plots are generated.   

```{r message=FALSE}
w.use <- wUseSample
areas <- c("Kent County","Sussex County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","TP.TotPop")
multi_element_data(w.use, data.elements, area.column = area.column,areas = areas)

```

Setting `plot.points=FALSE` will generate bar plots.

```{r message=FALSE}
w.use <- wUseSample
areas <- c("Kent County","Sussex County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","TP.TotPop")
multi_element_data(w.use, data.elements, plot.points = FALSE,
      area.column = area.column, areas = areas)

```

### Box Plots
Creates box plots for one or more elements. Multiple data elements are rendered as stacked plots (one plot per element).    

```{r message=FALSE}
w.use <- wUseSample
areas <- "15"
area.column = "STATECODE"
data.elements <- c("PS.GWPop","TP.TotPop")
boxplot_wu(w.use, data.elements, area.column = area.column,areas = areas)
```

If `areas = NA` then all areas are included in the plotted data. If `log = TRUE` the axis is changed to a log scale.

```{r message=FALSE}
w.use <- wUseSample
areas <- "15"
area.column = "STATECODE"
data.elements <- c("PS.GWPop","TP.TotPop")
boxplot_wu(w.use, data.elements, area.column, areas=NA, log=TRUE)

```

### Time series plots 
A time series plot shows water use for a data element for all areas of a subset of areas over time.  A user has the option to  illustration this using a bar graph or line graph, or point plot.  A selection of all areas or a subset of areas is also possible.  Only one data element per plot is possible but plots can be clustered to show multiple data elements in groups. 

  
```{r message=FALSE}
df <- wUseSample
areas <- c("Kent County","Sussex County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","TP.TotPop")
time_series_data(w.use, data.elements, area.column = area.column,areas = areas)
```

Setting `plot.points = FALSE` will change the format to a bar plot.

```{r message=FALSE}
df <- wUseSample
areas <- c("Kent County","Sussex County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","TP.TotPop")
time_series_data(w.use, data.elements, plot.points = FALSE, area.column = area.column,areas = areas)

```

If `years` are not specified, all years in the dataset are plotted. In the example below, years are specified.

```{r message=FALSE}
df <- wUseSample
areas <- c("Hawaii County","Honolulu County","Kauai County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","TP.TotPop")
time_series_data(w.use, data.elements, plot.points = FALSE,
       area.column = area.column, areas = areas, legend=TRUE, years = c("1995", "2000", "2005"))

```

The default range of the y-axis can be over-ridden by a user-specified range.

```{r message=FALSE}
df <- wUseSample
areas <- c("Hawaii County","Honolulu County","Kauai County")
area.column = "COUNTYNAME"
data.elements <- c("PS.GWPop","TP.TotPop")
time_series_data(w.use, data.elements, area.column, areas = areas, plot.points = FALSE, y.scale = c(0,1000))

```

### Bar Sums
This function is similar to the time series function as it creates bar plots for one or more data elements and years, however it will generate stacked bar plots comprising all data elements specified. Multiple stacked plots are generated if more than one area (e.g. county) is specified (one plot per area). 

```{r message=FALSE}
w.use <- wUseSample
areas <- c("New Castle County", "Kent County")
area.column = "COUNTYNAME"
data.elements <- c("PS.WTotl","CO.WTotl","DO.WTotl","IN.WTotl","PF.WTotl")
barchart_sums(w.use, data.elements, area.column = area.column,areas = areas)

```

The data elements can be unstacked by setting `plot.stack = FALSE`.

```{r message=FALSE}
w.use <- wUseSample
areas <- c("New Castle County", "Kent County")
area.column = "COUNTYNAME"
data.elements <- c("PS.WTotl","CO.WTotl","DO.WTotl","IN.WTotl","PF.WTotl")
barchart_sums(w.use, data.elements, plot.stack = FALSE, area.column = area.column,areas = areas)
```

The default range of the y-axis scale can be over-ridden by the user with `y.scale`.

```{r message=FALSE}
w.use <- wUseSample
areas <- c("New Castle County", "Kent County")
area.column = "COUNTYNAME"
data.elements <- c("PS.WTotl","CO.WTotl","DO.WTotl","IN.WTotl","PF.WTotl")
barchart_sums(w.use, data.elements, area.column, areas=areas, y.scale = c(0,500))
```

### Ranked data element tables
Users can rank data elements using the shiny application.  Once data is loaded, users can selected areas, years and data element types to display tables of values, which can be displayed in various ranking forms by clicking on the column name. 

### Choropleth
This function generates a map of counties shaded on a scaled color gradient representing the values of a data element and year of interest. Light shading corresponds to low values, dark shading corresponds to high values.

```{r message=FALSE}
w.use <- wUseSample
data.elements <- "PS.WFrTo"
year <- 2010 
areas <- "Hawaii" 
area.column <- "STATE_TERR"
ch.plot <- choropleth_plot(w.use, data.elements, year, areas, area.column)
```

The primary data element can be normalized to a secondary data element. The example below normalizes fresh water public supply by total population.

```{r message=FALSE}
w.use <- wUseSample
data.elements <- "PS.WFrTo"
norm.element <- "PS.TOPop"
year <- 2010 
areas <- "Hawaii" 
area.column <- "STATE_TERR"
ch.plot <- choropleth_plot(w.use, data.elements, year, areas, area.column, norm.element)
```

# Disclaimer
Software created by USGS employees along with contractors and grantees (unless specific stipulations are made in a contract or grant award) are to be released as Public Domain and free of copyright or license. Contributions of software components such as specific algorithms to existing software licensed through a third party are encouraged, but those contributions should be annotated as freely available in the Public Domain wherever possible. If USGS software uses existing licensed components, those licenses must be adhered to and redistributed.

Although this software has been used by the U.S. Geological Survey (USGS), no warranty, expressed or implied, is made by the USGS or the U.S. Government as to accuracy and functionality, nor shall the fact of distribution constitute any such warranty, and no responsibility is assumed by the USGS in connection therewith.